---
title: "Australian Multivariate Core Trend Inflation"
subtitle: "Bayesian dynamic-factor decomposition of monthly and quarterly CPI"
date: today
date-format: "MMMM YYYY"
---

```{r setup}
#| include: false
suppressMessages({
  library(targets); library(tarchetypes)
  library(dplyr); library(tibble); library(lubridate)
  library(ggplot2); library(scales)
  library(loo); library(posterior); library(arrow)
})

PROJECT_ROOT <- if (file.exists("_targets.R")) {
  normalizePath(".")
} else if (file.exists("../_targets.R")) {
  normalizePath("..")
} else {
  stop("Cannot locate _targets.R from ", getwd())
}
for (f in list.files(file.path(PROJECT_ROOT, "R"), "\\.R$", full.names = TRUE)) source(f)
targets::tar_config_set(store = file.path(PROJECT_ROOT, "_targets"))

tar_load(c(stan_data, weights, trend_A, common_share_A, sector_contribs_A, fit_A))
fit_A <- readRDS(fit_A)

latest_month <- max(stan_data$dates)
latest_trend <- tail(trend_A$median, 1)
latest_trend_ci <- c(tail(trend_A$lower, 1), tail(trend_A$upper, 1))
```

::: {.callout-note}
**Headline.** Trend inflation in `r format(latest_month, "%B %Y")` is
**`r sprintf("%.2f%%", latest_trend)`** (68% CI
`r sprintf("[%.2f, %.2f]", latest_trend_ci[1], latest_trend_ci[2])`) on the
demeaned, annualised scale used by this model. Values are differences from
each group's 2000–2019 mean inflation rate — add back the long-run mean to
read as headline % per year. The headline uses Variant A (AR(1) common
factor), which PSIS-LOO prefers over the Stock-Watson random-walk Variant B
and the combined Variant C on this sample. This is the **automated monthly
build** showing the Variant A panels only; the full three-variant
comparison is available in the manually-built dashboard.
:::

## Trend inflation

The headline output of the model — the AR(1)-common variant (Variant A),
preferred by LOO. Trend in Variant A is the weighted sum of sector-specific
trends $\sum_i w_i s_{i,t}$; the common factor is treated as a transitory
cycle. The band is a 68% posterior credible interval; the line is the
posterior median.

```{r trend-headline}
plot_trend_band(
  trend_A,
  title = NULL,
  subtitle = sprintf("Variant A (AR(1) common factor) · latest month: %s",
                     format(latest_month, "%B %Y")),
  caption = "Source: ABS CPI 6401.0 (Table 2 monthly from Apr-2024, Table 18 quarterly). Annualised, demeaned vs 2000-2019 sectoral means."
)
```

## How much of the move is *common*?

The share of weighted-aggregate inflation variance attributable to the common
factor (vs sector-specific idiosyncrasies). A reading above 50% means the
month's inflation move is broad-based; below 50% means it is sector-driven.

```{r common-share}
plot_common_share(
  common_share_A,
  subtitle = "Variant A — median + 68% credible band"
)
```

## Sector contributions to trend inflation (last 24 months)

Each band is sector $i$'s contribution $w_i \cdot s_{i,t}$ to the headline
trend. Stacking sums to (approximately) the headline trend median.

```{r sector-contribs, fig.height=5}
plot_sector_contribs(
  sector_contribs_A,
  since = latest_month %m-% months(23),
  subtitle = sprintf("Variant A · %s to %s",
                     format(latest_month %m-% months(23), "%b %Y"),
                     format(latest_month, "%b %Y"))
)
```

## Revisions vs prior vintage

How does this month's trend estimate compare to the previous monthly run at
the same dates? Solid line = current vintage (with band); dashed line =
prior.

```{r revisions}
#| fig-height: 3.5
prior_fit_path <- file.path(PROJECT_ROOT, "outputs/draws/fit_A_prior.rds")
prior_data_path <- file.path(PROJECT_ROOT, "outputs/draws/stan_data_prior.rds")

if (file.exists(prior_fit_path) && file.exists(prior_data_path)) {
  fit_A_prior <- readRDS(prior_fit_path)
  stan_data_prior <- readRDS(prior_data_path)
  rev <- posterior_revisions(fit_A, fit_A_prior,
                              stan_data$dates, stan_data_prior$dates)
  plot_revisions(rev,
                 subtitle = "Solid: current; dashed: prior month vintage")
} else {
  cat("_No prior vintage saved at `outputs/draws/fit_A_prior.rds`. The revisions panel becomes active after the next monthly run._")
}
```

### Which sectors drove the latest revision?

If a prior vintage is available, the bar chart below decomposes the change in
median trend at the most recent date into per-sector contributions. Bars to
the right of zero are sectors whose contribution to the headline grew between
vintages (pushing trend up); bars to the left pulled trend down.

```{r revision-attribution}
#| fig-height: 3.5
if (file.exists(prior_fit_path) && file.exists(prior_data_path)) {
  decomp <- decompose_revisions(
    fit_A, fit_A_prior,
    stan_data$dates, stan_data_prior$dates,
    weights = weights$weight,
    groups  = stan_data$groups
  )
  plot_revision_attribution(decomp, at_date = latest_month,
                            subtitle = format(latest_month, "%B %Y"))
} else {
  cat("_Attribution panel becomes active after the next monthly run._")
}
```

## Methodology (Variant A)

Let $\pi_{i,t}$ be the annualised, demeaned period-on-period inflation of CPI
group $i \in \{1,\dots,11\}$ in month $t$ on a latent monthly clock. The
measurement equation is

$$
\pi_{i,t} = \lambda_{i,t}\, c_t + s_{i,t}
            + \exp(h^\varepsilon_{i,t}/2)\,\varepsilon_{i,t}.
$$

A **monthly** observation at $t$ (April 2024 onward) equals $\pi_{i,t}$
directly. A **quarterly** observation placed at the last month of the
quarter equals the simple average of the three latent monthly inflations:

$$
y^Q_{i,t} = \tfrac{1}{3}\big(\pi_{i,t-2} + \pi_{i,t-1} + \pi_{i,t}\big),
$$

with measurement-noise variance averaged accordingly.

The common factor is modelled as a stationary AR(1) cycle:
$c_t = \rho\, c_{t-1} + \exp(h^c_t/2)\,\varepsilon^c_t$, $|\rho| < 1$.
Trend is $\tau_t = \sum_i w_i s_{i,t}$.

Estimated in Stan via `cmdstanr` with 4 chains × 1500 warmup + 1500 sampling
iterations and all random walks parameterised non-centrally.

---

::: {style="color:#666; font-size:0.85rem;"}
Automated monthly build (Variant A only). Full three-variant comparison and
LOO methodology section are in the manually-built dashboard. Built with
Stan, cmdstanr, targets, and Quarto.
:::
