---
title: "Australian Multivariate Core Trend Inflation"
subtitle: "A Stock-Watson decomposition of quarterly ABS CPI sub-groups"
date: today
date-format: "MMMM YYYY"
---

```{r setup}
#| include: false
suppressMessages({
  library(targets); library(tarchetypes)
  library(dplyr); library(tibble); library(lubridate)
  library(ggplot2); library(scales)
  library(loo); library(posterior); library(arrow); library(MASS)
})

PROJECT_ROOT <- if (file.exists("_targets.R")) {
  normalizePath(".")
} else if (file.exists("../_targets.R")) {
  normalizePath("..")
} else {
  stop("Cannot locate _targets.R from ", getwd())
}
for (f in list.files(file.path(PROJECT_ROOT, "R"),       "\\.R$", full.names = TRUE)) source(f)
for (f in list.files(file.path(PROJECT_ROOT, "R/gibbs"), "\\.R$", full.names = TRUE)) source(f)
targets::tar_config_set(store = file.path(PROJECT_ROOT, "_targets"))

tar_load(c(stan_data, weights, infl_demeaned,
           headline_yoy, trimmed_mean_yoy,
           trend_B, common_share_B, sector_contribs_B,
           fit_B, fit_B_wrap, loo_B))
fit_B <- readRDS(fit_B)

# Convert the model's demeaned trend back to raw % so it's on the same
# scale as headline CPI YoY (the dashboard overlay).
offset <- demean_offset(infl_demeaned, weights)
trend_raw <- trend_B
trend_raw$median <- trend_raw$median + offset
trend_raw$lower  <- trend_raw$lower  + offset
trend_raw$upper  <- trend_raw$upper  + offset

latest_quarter <- max(stan_data$dates)
latest_trend   <- tail(trend_raw$median, 1)
latest_trend_ci <- c(tail(trend_raw$lower, 1), tail(trend_raw$upper, 1))
latest_cs      <- tail(common_share_B$median, 1)
latest_cs_ci   <- c(tail(common_share_B$lower, 1),
                    tail(common_share_B$upper, 1))
```

::: {.callout-note appearance="minimal" icon="false"}
### Multivariate Core Trend &nbsp;·&nbsp; `r format(latest_quarter, "%B %Y")`

**`r sprintf("%.2f%%", latest_trend)`** per year &nbsp;·&nbsp;
68% credible interval `r sprintf("%.2f–%.2f%%", latest_trend_ci[1], latest_trend_ci[2])`

A model-based measure of underlying Australian CPI inflation, applying
the [NY Fed Multivariate Core Trend](https://www.newyorkfed.org/research/policy/mct)
methodology to 33 seasonally-adjusted ABS CPI sub-group series.
Estimates the broad-based, persistent component of the rate of price
change, stripped of transitory shocks and sector-specific noise.
Re-estimated each quarter.
:::

## Trend inflation

The model's estimate of underlying inflation — median + 68% posterior
credible band — plotted alongside headline CPI inflation (year-on-year)
for context.

```{r trend-headline}
plot_trend_band(
  trend_raw,
  title    = NULL,
  subtitle = sprintf("Annualised, %% per year. Latest: %s",
                     format(latest_quarter, "%B %Y")),
  caption  = "Source: ABS CPI 6401.0; author's calculations.",
  ylab     = "Percent",
  overlays = list(
    "Headline CPI (YoY)"     = dplyr::rename(headline_yoy,     value = "yoy"),
    "Trimmed mean CPI (YoY)" = dplyr::rename(trimmed_mean_yoy, value = "yoy")
  ),
  since    = latest_quarter %m-% years(9)
)
```

## Group contributions to trend inflation

The 33 sub-group trend contributions $w_i \cdot s_{i,t}$ aggregated back
to the 11 standard ABS CPI groups. The bar chart ranks each group's
contribution to the most recent quarter; the small-multiples panel
below traces every group's contribution across the full sample.

```{r contribs-data}
parent_contribs <- aggregate_contribs_by_parent(
  sector_contribs_B,
  parent_map   = abs_subgroup_parent(),
  parent_order = abs_cpi_groups()
)
```

```{r contribs-bar, fig.height=3.5}
plot_contribs_bar(
  parent_contribs, at_date = latest_quarter,
  subtitle = format(latest_quarter, "%B %Y"),
  caption  = "Source: ABS CPI 6401.0 Table 18; author's calculations."
)
```

```{r contribs-facet, fig.height=7}
plot_sector_contribs(
  parent_contribs,
  subtitle = sprintf("%s to %s",
                     format(min(parent_contribs$date), "%b %Y"),
                     format(latest_quarter, "%b %Y")),
  caption  = "Source: ABS CPI 6401.0 Table 18; author's calculations."
)
```

```{r revisions-section}
#| results: asis
#| fig-height: 3.5
prior_fit_path  <- file.path(PROJECT_ROOT, "outputs/draws/fit_B_prior.rds")
prior_data_path <- file.path(PROJECT_ROOT, "outputs/draws/stan_data_prior.rds")
has_prior <- file.exists(prior_fit_path) && file.exists(prior_data_path)

if (has_prior) {
  fit_B_prior_raw  <- readRDS(prior_fit_path)
  stan_data_prior  <- readRDS(prior_data_path)
  fit_B_prior_wrap <- mct_gibbs_fit(fit_B_prior_raw, stan_data_prior)

  cat("## Revisions vs prior vintage\n\n",
      "How does this quarter's trend estimate compare to the previous run ",
      "at the same dates? Solid line = current vintage (with band); ",
      "dashed line = prior.\n\n", sep = "")
  rev <- posterior_revisions(fit_B_wrap, fit_B_prior_wrap,
                             stan_data$dates, stan_data_prior$dates)
  print(plot_revisions(rev,
                       subtitle = "Solid: current; dashed: prior vintage"))

  cat("\n\n### Which sub-groups drove the latest revision?\n\n",
      "Bars to the right of zero are sub-groups whose contribution to the ",
      "headline grew between vintages (pushing trend up); bars to the left ",
      "pulled trend down.\n\n", sep = "")
  decomp <- decompose_revisions(
    fit_B_wrap, fit_B_prior_wrap,
    stan_data$dates, stan_data_prior$dates,
    weights = weights$weight,
    groups  = stan_data$groups
  )
  print(plot_revision_attribution(decomp, at_date = latest_quarter,
                                  subtitle = format(latest_quarter, "%B %Y")))
}
```

## Sector decomposition: common vs sector-specific

The trend is the weighted aggregate $\tau_t = \sum_i w_i \big( \lambda_i
c_t + s_{i,t} \big)$. Splitting that into a **common** part (broad-based
drift from the common trend, $\lambda_i c_t$, multiplied by each
sub-group's weight) and a **sector-specific** part (each sub-group's own
random-walk drift $s_{i,t}$), then aggregating to three macro buckets
(Goods / Services ex. housing / Housing). Bands above zero pushed trend
up; below zero pulled it down.

```{r decomposition, fig.height=5}
decomp <- decompose_trend_by_bucket(
  fit_B_wrap, weights = weights$weight,
  dates = stan_data$dates, sub_groups = stan_data$groups,
  bucket_map = abs_subgroup_to_macrobucket()
)
plot_sector_decomposition(
  decomp = decomp,
  trend  = trend_B,
  subtitle = sprintf("%s to %s",
                     format(latest_quarter %m-% years(9), "%b %Y"),
                     format(latest_quarter, "%b %Y")),
  caption  = "Source: ABS CPI 6401.0; author's calculations.",
  since    = latest_quarter %m-% years(9)
)
```

## How much of the move is *common*?

The share of weighted-aggregate inflation variance attributable to the
common trend (vs sub-group-specific idiosyncrasies). For AU sub-group
inflation this is currently `r sprintf("%.0f%%", 100 * latest_cs)`
(68% CI `r sprintf("[%.0f%%, %.0f%%]", 100 * latest_cs_ci[1], 100 * latest_cs_ci[2])`)
— most inflation moves are broad-based across sub-groups, not driven by
any single category. The y-axis is zoomed to the posterior range so
period-to-period changes are visible; the absolute level has been high
(>80%) for the entire sample.

```{r common-share}
plot_common_share(
  common_share_B,
  subtitle = "Median + 68% credible band",
  caption  = "Source: ABS CPI 6401.0 Table 18; author's calculations."
)
```

## Methodology

Let $\pi_{i,t}$ be the annualised, demeaned quarter-on-quarter inflation of
ABS CPI sub-group $i \in \{1,\dots,33\}$ in quarter $t$. The measurement
equation is

$$
\pi_{i,t} = \lambda_i\, c_t + s_{i,t}
            + \exp(h^\varepsilon_{i,t}/2)\,
              s_{i,t}^{\text{out}}\,
              \big(u_{i,t} + \theta_i u_{i,t-1}\big),
$$

where $\lambda_i$ is a sub-group-specific loading on the common trend
($\lambda_{\text{Rents}} = 1$ pins identification), $u_{i,t}
\sim \mathcal{N}(0,1)$, and $\theta_i$ is a sub-group-specific MA(1)
coefficient that lets a shock to sector $i$ bleed out into the next
quarter. $s_{i,t}^{\text{out}}$ is a per-observation scale-mixture
indicator: most periods take $s = 1$ (normal observation) but a Bernoulli
gate $\Pr(s = 1) = p_i$ allows the model to flag anomalous-volatility
quarters and discount their influence on the trend. State equations are
random walks with stochastic volatility:

$$
c_t = c_{t-1} + \exp(h^c_t/2)\,\eta^c_t, \qquad
s_{i,t} = s_{i,t-1} + \exp(h^s_{i,t}/2)\,\eta^s_{i,t}.
$$

Log-volatilities $h^c_t$, $h^s_{i,t}$, $h^\varepsilon_{i,t}$ are themselves
random walks. Trend is the weighted aggregate
$\tau_t = \sum_i w_i\big(\lambda_i c_t + s_{i,t}\big)$ with weights from
the ABS 2025 Weighting Pattern (sub-group level, sums to 1).

The model specification follows the [NY Fed Multivariate Core Trend
(MCT)](https://www.newyorkfed.org/research/policy/mct) — see Almuzara
and Sbordone, "[Inflation Persistence: How Much Is There and Where Is
It Coming From?](https://libertystreeteconomics.newyorkfed.org/2022/04/inflation-persistence-how-much-is-there-and-where-is-it-coming-from/)",
Liberty Street Economics, April 2022. This dashboard applies that
methodology to Australian CPI data: 33 ABS sub-groups (seasonally
adjusted) at quarterly frequency, with an MA(1) sector-noise component
and a scale-mixture outlier mechanism. Estimated by Gibbs sampling
(a custom R port of the NY Fed MATLAB sampler) on the
inflation-targeting-era sample (`r format(min(stan_data$dates), "%Y-Q%q")`
to `r format(max(stan_data$dates), "%Y-Q%q")`,
T = `r stan_data$T` quarters, N = 33 sub-groups).

---

::: {style="color:#666; font-size:0.85rem;"}
Built with R, targets and Quarto. Re-runs automatically each quarter on
the publication day of the ABS CPI release.
:::
