---
title: "Australian Multivariate Core Trend Inflation"
subtitle: "Bayesian dynamic-factor decomposition of monthly and quarterly CPI"
date: today
date-format: "MMMM YYYY"
---

```{r setup}
#| include: false
suppressMessages({
  library(targets); library(tarchetypes)
  library(dplyr); library(tibble); library(lubridate)
  library(ggplot2); library(scales)
  library(loo); library(posterior); library(arrow)
})

# Project source files. When rendered via `tar_quarto()`, the working
# directory is `dashboard/`, so paths step up one level.
for (f in list.files("../R", "\\.R$", full.names = TRUE)) source(f)

tar_load(c(
  stan_data, weights,
  trend_A, trend_B, trend_C,
  common_share_A, common_share_B, common_share_C,
  sector_contribs_A, sector_contribs_B, sector_contribs_C,
  fit_A, fit_B, fit_C, variant_comparison
))

latest_month <- max(stan_data$dates)
latest_trend <- tail(trend_A$median, 1)
latest_trend_ci <- c(tail(trend_A$lower, 1), tail(trend_A$upper, 1))
```

::: {.callout-note}
**Headline.** Trend inflation in `r format(latest_month, "%B %Y")` is
**`r sprintf("%.2f%%", latest_trend)`** (68% CI
`r sprintf("[%.2f, %.2f]", latest_trend_ci[1], latest_trend_ci[2])`) on the
demeaned, annualised scale used by this model. Values are differences from
each group's 2000–2019 mean inflation rate — add back the long-run mean to
read as headline % per year. The headline uses Variant A (AR(1) common
factor), which PSIS-LOO prefers over the Stock-Watson random-walk Variant B
on this sample — see the variant-comparison section below.
:::

## Trend inflation

The headline output of the model — the AR(1)-common variant (Variant A),
preferred by LOO. Trend in Variant A is the weighted sum of sector-specific
trends $\sum_i w_i s_{i,t}$; the common factor is treated as a transitory
cycle. The band is a 68% posterior credible interval; the line is the
posterior median.

```{r trend-headline}
plot_trend_band(
  trend_A,
  title = NULL,
  subtitle = sprintf("Variant A (AR(1) common factor) · latest month: %s",
                     format(latest_month, "%B %Y")),
  caption = "Source: ABS CPI 6401.0 (Table 2 monthly from Apr-2024, Table 18 quarterly). Annualised, demeaned vs 2000-2019 sectoral means."
)
```

## How much of the move is *common*?

The share of weighted-aggregate inflation variance attributable to the common
factor (vs sector-specific idiosyncrasies). A reading above 50% means the
month's inflation move is broad-based; below 50% means it is sector-driven.
See the caveat in the methodology section about Variant A's common-share
metric — the apples-to-apples Stock-Watson reading is Variant B's, plotted
in the variant comparison below.

```{r common-share}
plot_common_share(
  common_share_A,
  subtitle = "Variant A — median + 68% credible band"
)
```

## Sector contributions to trend inflation (last 24 months)

Each band is sector $i$'s contribution $w_i \cdot s_{i,t}$ to the headline
trend. Stacking sums to (approximately) the headline trend median.

```{r sector-contribs, fig.height=5}
plot_sector_contribs(
  sector_contribs_A,
  since = latest_month %m-% months(23),
  subtitle = sprintf("Variant A · %s to %s",
                     format(latest_month %m-% months(23), "%b %Y"),
                     format(latest_month, "%b %Y"))
)
```

## Revisions vs prior vintage

How does this month's trend estimate compare to the previous monthly run at
the same dates? Solid line = current vintage (with band); dashed line =
prior.

```{r revisions}
#| fig-height: 3.5
prior_fit_path <- "../outputs/draws/fit_A_prior.rds"
prior_data_path <- "../outputs/draws/stan_data_prior.rds"

if (file.exists(prior_fit_path) && file.exists(prior_data_path)) {
  fit_A_prior <- readRDS(prior_fit_path)
  stan_data_prior <- readRDS(prior_data_path)
  rev <- posterior_revisions(fit_A, fit_A_prior,
                              stan_data$dates, stan_data_prior$dates)
  plot_revisions(rev,
                 subtitle = "Solid: current; dashed: prior month vintage")
} else {
  cat("_No prior vintage saved at `outputs/draws/fit_A_prior.rds`. The revisions panel becomes active after the next monthly run — see the README for the rotation step._")
}
```

### Which sectors drove the latest revision?

If a prior vintage is available, the bar chart below decomposes the change in
median trend at the most recent date into per-sector contributions. Bars to
the right of zero are sectors whose contribution to the headline grew between
vintages (pushing trend up); bars to the left pulled trend down. This is the
MCMC-native analogue of the NY Fed MCT's Kalman-update attribution.

```{r revision-attribution}
#| fig-height: 3.5
if (file.exists(prior_fit_path) && file.exists(prior_data_path)) {
  decomp <- decompose_revisions(
    fit_A, fit_A_prior,
    stan_data$dates, stan_data_prior$dates,
    weights = weights$weight,
    groups  = stan_data$groups
  )
  plot_revision_attribution(decomp, at_date = latest_month,
                            subtitle = format(latest_month, "%B %Y"))
} else {
  cat("_Attribution panel becomes active after the next monthly run._")
}
```

## Variant comparison

Three specifications of the common factor are estimated. **Variant A** treats
the common factor as a stationary AR(1) cycle (its trend is the weighted
sum of sector-specific drifts). **Variant B** treats it as a random walk —
the canonical Stock-Watson common trend. **Variant C** combines both —
a random-walk common trend AND an i.i.d. common cycle, each with its own
stochastic volatility — nesting A and B as restrictions. On this sample
LOO prefers A by a wide margin over C, indicating Australian sectoral
co-movement is overwhelmingly transitory rather than trend-driven.

```{r variant-trend-compare}
compare_trend_plot(fit_A, fit_B, fit_C, dates = stan_data$dates)
```

```{r variant-cs-compare}
compare_common_share_plot(fit_A, fit_B, fit_C, dates = stan_data$dates)
```

`r variant_comparison$text`

```{r loo-table}
#| echo: false
loo_df <- as.data.frame(variant_comparison$loo_compare)
loo_df$model <- rownames(loo_df)
loo_df <- loo_df[, c("model", "elpd_diff", "se_diff", "elpd_loo", "se_elpd_loo",
                     "p_loo", "se_p_loo")]
knitr::kable(loo_df, digits = 1, row.names = FALSE,
             caption = "LOO comparison (elpd, with standard errors).")
```

## Methodology

Let $\pi_{i,t}$ be the annualised, demeaned period-on-period inflation of CPI
group $i \in \{1,\dots,11\}$ in month $t$ on a latent monthly clock. The
measurement equation is

$$
\pi_{i,t} = \lambda_{i,t}\, c_t + s_{i,t}
            + \exp(h^\varepsilon_{i,t}/2)\,\varepsilon_{i,t}.
$$

A **monthly** observation at $t$ (April 2024 onward) equals $\pi_{i,t}$
directly. A **quarterly** observation placed at the last month of the
quarter equals the simple average of the three latent monthly inflations:

$$
y^Q_{i,t} = \tfrac{1}{3}\big(\pi_{i,t-2} + \pi_{i,t-1} + \pi_{i,t}\big),
$$

with measurement-noise variance averaged accordingly.

The states evolve as

$$
\begin{aligned}
\lambda_{i,t} &= \lambda_{i,t-1} + \sigma_\lambda\, \eta^\lambda_{i,t}, \\
s_{i,t} &= s_{i,t-1} + \exp(h^s_{i,t}/2)\, \eta^s_{i,t}, \\
h^c_t &= h^c_{t-1} + \sigma_{h_c}\, \eta^{h_c}_t,
\end{aligned}
$$

with $h^s_{i,t}, h^\varepsilon_{i,t}$ analogous. The reference loading
$\lambda_{\text{Housing},t} = 1$ for all $t$ identifies the common factor's
sign and scale.

**Variant A** (AR(1) common):
$c_t = \rho\, c_{t-1} + \exp(h^c_t/2)\,\varepsilon^c_t$, $|\rho| < 1$;
trend is $\tau_t = \sum_i w_i s_{i,t}$.

**Variant B** (RW common trend):
$c_t = c_{t-1} + \exp(h^c_t/2)\,\eta^c_t$;
trend is $\tau_t = (\sum_i w_i \lambda_{i,t})\, c_t + \sum_i w_i s_{i,t}$.

**Variant C** (combined trend + cycle):
$\tau^c_t = \tau^c_{t-1} + \exp(h^c_t/2)\,\eta^{c}_t$ (random-walk common
trend) and $c_t = \exp(h^{ec}_t/2)\,\varepsilon^c_t$ (i.i.d. common cycle),
with sector-specific loadings $\lambda^\tau_i$ on the trend and
$\lambda^\varepsilon_i$ on the cycle (both time-invariant; Housing pinned
at 1 for identification). The measurement equation becomes $\pi_{i,t} =
\lambda^\tau_i \tau^c_t + \lambda^\varepsilon_i c_t + s_{i,t} +
\exp(h^\varepsilon_{i,t}/2)\,\varepsilon_{i,t}$; trend is $\tau_t =
\sum_i w_i (\lambda^\tau_i \tau^c_t + s_{i,t})$. Variant C nests A
($\tau^c = 0$) and B ($c = 0$) as restrictions.

Estimated in Stan via `cmdstanr` with all random walks parameterised
non-centrally. Weights $w_i$ are the 2025 ABS Weighting Pattern, fixed
across the sample. A and B use 4 chains × 1500 warmup + 1500 sampling
iterations; C uses 6 chains × 1500 + 1500 because its two-common-factor
structure is at the edge of HMC mixing reliability on this sample size.

### Caveats

- `common_share` in Variants A and C compares common-INNOVATION variance
  to sector-TREND innovation variance — an awkward apples-to-oranges
  metric forced into them for cross-variant comparability. The canonical
  Stock-Watson interpretation only applies to Variant B. (For C it
  measures the contribution of the *common-trend* component $\tau^c$ only;
  the common cycle $c$ does not enter the variance ratio.)
- The model uses fixed expenditure weights from the latest vintage. Real
  weight shifts (e.g., the 2017 reweighting moving methodology to annual
  HFCE-based updates) are not reflected.
- PSIS-LOO Pareto-k diagnostics for state-space models often flag many high
  values; treat the LOO ranking as indicative.

---

::: {style="color:#666; font-size:0.85rem;"}
Built with Stan, cmdstanr, targets, and Quarto. Source:
[github.com/...] · Methodology and code in the project's `CLAUDE.md` brief.
:::
