---
title: "Australian Multivariate Core Trend Inflation"
subtitle: "Bayesian dynamic-factor decomposition of monthly and quarterly CPI"
date: today
date-format: "MMMM YYYY"
---

```{r setup}
#| include: false
suppressMessages({
  library(targets); library(tarchetypes)
  library(dplyr); library(tibble); library(lubridate)
  library(ggplot2); library(scales)
  library(loo); library(posterior); library(arrow)
})

# Project source files. When rendered via `tar_quarto()`, the working
# directory is `dashboard/`, so paths step up one level.
for (f in list.files("../R", "\\.R$", full.names = TRUE)) source(f)

tar_load(c(
  stan_data, weights,
  trend_A, trend_B, common_share_A, common_share_B,
  sector_contribs_A, sector_contribs_B,
  fit_A, fit_B, variant_comparison
))

latest_month <- max(stan_data$dates)
latest_trend <- tail(trend_B$median, 1)
latest_trend_ci <- c(tail(trend_B$lower, 1), tail(trend_B$upper, 1))
```

::: {.callout-note}
**Headline.** Trend inflation in `r format(latest_month, "%B %Y")` is
**`r sprintf("%.2f%%", latest_trend)`** (68% CI
`r sprintf("[%.2f, %.2f]", latest_trend_ci[1], latest_trend_ci[2])`) on the
demeaned, annualised scale used by this model. Values are differences from
each group's 2000–2019 mean inflation rate — add back the long-run mean to
read as headline % per year.
:::

## Trend inflation

The headline output of the model — the common-trend variant (Variant B). The
band is a 68% posterior credible interval; the line is the posterior median.

```{r trend-headline}
plot_trend_band(
  trend_B,
  title = NULL,
  subtitle = sprintf("Variant B (random-walk common trend) · latest month: %s",
                     format(latest_month, "%B %Y")),
  caption = "Source: ABS CPI 6401.0 (Table 2 monthly from Apr-2024, Table 18 quarterly). Annualised, demeaned vs 2000-2019 sectoral means."
)
```

## How much of the move is *common*?

The share of weighted-aggregate inflation variance attributable to the common
factor (vs sector-specific idiosyncrasies). A reading above 50% means the
month's inflation move is broad-based; below 50% means it is sector-driven.

```{r common-share}
plot_common_share(
  common_share_B,
  subtitle = "Variant B — median + 68% credible band"
)
```

## Sector contributions to trend inflation (last 24 months)

Each band is sector $i$'s contribution $w_i \cdot s_{i,t}$ to the headline
trend. Stacking sums to (approximately) the headline trend median.

```{r sector-contribs, fig.height=5}
plot_sector_contribs(
  sector_contribs_B,
  since = latest_month %m-% months(23),
  subtitle = sprintf("Variant B · %s to %s",
                     format(latest_month %m-% months(23), "%b %Y"),
                     format(latest_month, "%b %Y"))
)
```

## Revisions vs prior vintage

How does this month's trend estimate compare to the previous monthly run at
the same dates? Solid line = current vintage (with band); dashed line =
prior.

```{r revisions}
#| fig-height: 3.5
prior_fit_path <- "../outputs/draws/fit_B_prior.rds"
prior_data_path <- "../outputs/draws/stan_data_prior.rds"

if (file.exists(prior_fit_path) && file.exists(prior_data_path)) {
  fit_B_prior <- readRDS(prior_fit_path)
  stan_data_prior <- readRDS(prior_data_path)
  rev <- posterior_revisions(fit_B, fit_B_prior,
                              stan_data$dates, stan_data_prior$dates)
  plot_revisions(rev,
                 subtitle = "Solid: current; dashed: prior month vintage")
} else {
  cat("_No prior vintage saved at `outputs/draws/fit_B_prior.rds`. The revisions panel becomes active after the next monthly run — see the README for the rotation step._")
}
```

## Variant comparison

Two specifications of the common factor are estimated. Variant A treats the
common factor as transitory (AR(1)); Variant B treats it as a random walk —
the canonical Stock-Watson trend.

```{r variant-trend-compare}
compare_trend_plot(fit_A, fit_B, stan_data$dates)
```

```{r variant-cs-compare}
compare_common_share_plot(fit_A, fit_B, stan_data$dates)
```

`r variant_comparison$text`

```{r loo-table}
#| echo: false
loo_df <- as.data.frame(variant_comparison$loo_compare)
loo_df$model <- rownames(loo_df)
loo_df <- loo_df[, c("model", "elpd_diff", "se_diff", "elpd_loo", "se_elpd_loo",
                     "p_loo", "se_p_loo")]
knitr::kable(loo_df, digits = 1, row.names = FALSE,
             caption = "LOO comparison (elpd, with standard errors).")
```

## Methodology

Let $\pi_{i,t}$ be the annualised, demeaned period-on-period inflation of CPI
group $i \in \{1,\dots,11\}$ in month $t$ on a latent monthly clock. The
measurement equation is

$$
\pi_{i,t} = \lambda_{i,t}\, c_t + s_{i,t}
            + \exp(h^\varepsilon_{i,t}/2)\,\varepsilon_{i,t}.
$$

A **monthly** observation at $t$ (April 2024 onward) equals $\pi_{i,t}$
directly. A **quarterly** observation placed at the last month of the
quarter equals the simple average of the three latent monthly inflations:

$$
y^Q_{i,t} = \tfrac{1}{3}\big(\pi_{i,t-2} + \pi_{i,t-1} + \pi_{i,t}\big),
$$

with measurement-noise variance averaged accordingly.

The states evolve as

$$
\begin{aligned}
\lambda_{i,t} &= \lambda_{i,t-1} + \sigma_\lambda\, \eta^\lambda_{i,t}, \\
s_{i,t} &= s_{i,t-1} + \exp(h^s_{i,t}/2)\, \eta^s_{i,t}, \\
h^c_t &= h^c_{t-1} + \sigma_{h_c}\, \eta^{h_c}_t,
\end{aligned}
$$

with $h^s_{i,t}, h^\varepsilon_{i,t}$ analogous. The reference loading
$\lambda_{\text{Housing},t} = 1$ for all $t$ identifies the common factor's
sign and scale.

**Variant A** (AR(1) common):
$c_t = \rho\, c_{t-1} + \exp(h^c_t/2)\,\varepsilon^c_t$, $|\rho| < 1$;
trend is $\tau_t = \sum_i w_i s_{i,t}$.

**Variant B** (RW common trend):
$c_t = c_{t-1} + \exp(h^c_t/2)\,\eta^c_t$;
trend is $\tau_t = (\sum_i w_i \lambda_{i,t})\, c_t + \sum_i w_i s_{i,t}$.

Estimated in Stan via `cmdstanr` with all random walks parameterised
non-centrally. Weights $w_i$ are the 2025 ABS Weighting Pattern, fixed
across the sample. Inference uses 4 chains × 1500 warmup + 1500 sampling
iterations.

### Caveats

- `common_share` in Variant A compares common-INNOVATION variance to
  sector-TREND innovation variance — an awkward apples-to-oranges metric
  forced into Variant A for cross-variant comparability. The canonical
  Stock-Watson interpretation only applies to Variant B.
- The model uses fixed expenditure weights from the latest vintage. Real
  weight shifts (e.g., the 2017 reweighting moving methodology to annual
  HFCE-based updates) are not reflected.
- PSIS-LOO Pareto-k diagnostics for state-space models often flag many high
  values; treat the LOO ranking as indicative.

---

::: {style="color:#666; font-size:0.85rem;"}
Built with Stan, cmdstanr, targets, and Quarto. Source:
[github.com/...] · Methodology and code in the project's `CLAUDE.md` brief.
:::
